scRNAseq benchmark in want of a title

### 2020-01-22 bmc_article.tex

The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs]. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of coexpression and binary classification. \texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores which must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations. Estimates of variance explained by subpopulations based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the clustering accuracy (Figure \ref{fig:figure6}A). To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all dataset. The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and un-standardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

We next investigated the impact of the number of dimensions used on the separability of the subpopulations. Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact. We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA). As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition. We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields. Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method. Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization. Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e. maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations, suggesting that these components help classification despite being individually uninformative.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

>>>>

The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs]. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of co-expression and binary classification. \texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the mean of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3). In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. Estimates of variance explained by subpopulation based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulation, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulation (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the clustering accuracy (Figure \ref{fig:figure6}A). To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

We next investigated the impact of the number of dimensions used on the separability of the subpopulations. Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact. We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA). As a first approximation of the real dimension, we computed the variance in each principal component that was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition. We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields. Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method. Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization. Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e. maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations, suggesting that these components help classification despite being individually uninformative.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
